\plaintitle

### 2017-01-11 drafts/labor/piecework_lit.tex
many minor changes through complexity

\topic{The HCI community has used the term ``piecework'' to describe
myriad instantiations of on--demand labor, but
researchers have generally made this allusion in passing.}
Since this paper traces a much stronger parallel between
(historical) piecework and (contemporary) on--demand work,
a more comprehensive background on piecework will be useful.
Specifically,
first, we'll define ``piecework'' as researchers in its field understand it;
and second, we'll trace the rise and fall of piecework at a high level,
identifying key figures and ideas during this time.
This section is not intended to be comprehensive:
instead, it sets up the scaffolding necessary for
our later investigations of on--demand work's three questions:
complexity limits,
task decomposition, and
worker relationships.

@@

\topic{Aligning on--demand work with piecework requires an understanding of what piecework is.}
While it has had several definitions over the years,
we can trace a constellation of characteristics that recur throughout the literature.
We'll follow this research, collecting
descriptions,
examples, and
definitions,
to develop a working understanding.

\topic{Piecework's history traces back further perhaps than most would expect.}
\citeauthor{grier2013computers} describes the process astronomers adopted of hiring young boys
to calculate equations in order
to better--predict the trajectories of various celestial bodies in the 1830s~\cite{grier2013computers}.
George Airy was perhaps the first to rigorously apply piecework--style decomposition of tasks to work;
by breaking complex calculations into constituent parts, and
training young men to solve simple algebraic problems,
Airy could distribute work to many more people than could otherwise complete the full calculations.

\topic{Piecework may have started in the intellectual domain of astronomical calculations and projections,
but it found its foothold in manual labor.}
Piecework took hold in farm work~\cite{hughRaynbirdTaskWork},
in textiles~\cite{restructuringPieceworkBaker,riisOtherSideLives},
on railroads~\cite{Brown01041990}, and 
elsewhere in manufacturing~\cite{10.2307/3827491}.
At the turn of the \nth{20} century, \citeauthor{taylor1914principles}
--- a mechanical engineer with a fascination in work efficiency ---
began studying and formalizing the decomposition, tracking, and management of tasks.
In \citeyear{taylor1914principles} he published \citetitle{taylor1914principles},
concretizing an idea that had nebulously been forming,
and which he had been working out himself, for years~\cite{taylor1914principles}.
Scientific Management (and Fordism) thrust piecework into higher gear, especially as
mass manufacturing and
a depleted wartime workforce forced industry to find new ways to eke out more production capacity.

@@

\topic{By \citeyear{hughRaynbirdTaskWork} we find
a concise definition of piecework
in \citeauthor{hughRaynbirdTaskWork}'s essay on piecework,
particularly driven toward encapsulating the manual labor of farmwork.}
He does this by contrasting with the ``day--labourer'' ---
``the chief difference lies between the day--labourer,
who receives a certain some of money~\dots~for his day's work,
and the task--labourer, whose earnings depend on the quantity of work done''~\cite{hughRaynbirdTaskWork}.
\citeauthor{10.2307/2338394} defines it through examples:
``payment is made for each hectare which is pronounced to be well ploughed~\dots~for each living foal got from a mare;~\dots~for each living calf got''~\cite{10.2307/2338394}.
This framing offers an intuitive sense of piecework;
``payment for results,'' as he calls it,
is not only common in practice, but
well--studied in labor economics as well~\cite{Figlio2007901,weitzman1976new,10.2307/3003414,BJIR:BJIR038}.

\topic{It's worth acknowledging that
``this distinction [between piece--rates and time--rates] was not completely clear--cut''~\cite{hart2013rise}.}
Employers implemented piece--rates in some aspects and
time--rates in others.
The Rowan premium system, for example,
essentially paid workers
a base rate for time plus
additional pay dependent on output~\cite{rowan1901premium}.
As \citeauthor{rowan1901premium}'s premium system guaranteed an hourly rate
regardless of the worker's productive output
\textit{as well as} additional compensation tied to performance,
workers were
in some senses ``task--labourers'', but
in other senses ``day--labourers''.
This was just one of several alternatives to strict time-- and piece--rate renumeration paradigms.

@@

\topic{We see this sense of independence in myriad times, locales, and industries.}
\citeauthor{10.2307/3827491} offers a look into the lives and culture of ``match--girls''
--- teenage women who assembled matchsticks in the late \nth{19} century in London. % added that these women were in London since Chinmay critiqued that we're glossing over the labor advocacy movement that grew in Europe, which revealed that we completely left out the fact that this all happened in England.
\ali{Chinmay: but still not convinced; this is a post hoc ergo propter hoc argument}
Of interest was their reputation ``\dots~for generosity, independence, and protectiveness,
but also for brashness, irregularity, low morality, and little education''~\cite{10.2307/3827491}.
\citeauthor{10.2307/27508091} document piecework from 1850 through 1930 in Australia,
finding similar notions of independence and autonomy among piecework newspaper compositors:
``If a piece--work compositor~\dots~decided that he did not want to work on a particular day or night, the management recognised his right to put a `substitute' or `grass' compositor in his place''~\cite{10.2307/27508091}.
This sense of independence and autonomy appears to be a common component of piecework.

\topic{Since workers could now choose their own schedule and style, there arose a discussion on how best to manage pieceworkers.}
This conversation generally regarded workers antagonistically~\cite{roy1954efficiency}, a far cry from the earlier rhetoric on piecework, which promised that
piece workers would gladly work as diligently and as hard as possible because
incentive--based pay rewards hard work~\cite{clark1908cotton}.

@@

\topic{Piecework's popularity in the United States and Europe fell almost as quickly as it had climbed.}
Between 1938 and 1942, the proportion of metal workers under piecework systems had climbed steeply from 11\% to 60\%~\cite{hart2005piecework}.
By 1961, that proportion dropped to 8\%~\cite{carlson1982time}.
\citeauthor{carlson1982time} details that, from 1973 to 1980, the holdouts of piecework
--- where more than 50\% worked under incentive wage plans ---
were principally in clothes--making (e.g. hosiery, footwear, and garments).
\citeauthor{hart2013rise} offer a number of explanations for the sudden vanishing of piecework.
The salient suggestions include:
\begin{inlinelist}
\item the emergence of more effective, more nuanced incentive models
--- rewarding teams for complex achievements, for instance;
\item the shifting of piecework industries such as manufacturing and textiles to other countries; and
\item the quality of ``multidimensional'' work becoming too difficult to evaluate
\end{inlinelist}~\cite{hart2013rise}.

@@

Viewing on--demand work as a modern instantiation of piecework is relatively straightforward by this definition.
First, platforms such as Mechanical Turk, Uber and TaskRabbit pay by the task, though some such as Upwork do offer hourly rates as well.
Second, workers are attracted to these platforms by the freedom they offer to pick the time and place of work~\cite{martin2014being,whyWouldAnyoneBrewer}.
Third, system developers as on Mechanical Turk typically assume no professional skills in transcription or other areas, and attempt to build that expertise into the work flow~\cite{noronha2011platemate,bernsteinSoylent}.
Given this alignment, many of the same properties of piecework historically will apply to on--demand work as well.

\subsection{Case studies in piecework}
\topic{Throughout the rest of the paper, we will return to four case studies to frame our analyses:
Airy's employment of human computers;
domestic and farm workers;
the ``match--girls'' strike;
and industrial and assembly-line workers.}
In introducing these cases at a high level,
we'll trace the history of piecework
while also framing the later analysis of the major research threads we named earlier: complexity, decomposition, and relationships.

@@

This practice ensured two outcomes that disfavored workers.
First, it eliminated any potential to advance professionally, as
workers' careers ended relatively early in their careers,
and without formal education in mathematics they struggled to find work for which their experience was meaningful.
Second, it limited workers' ability to organize
by ensuring that workers were barely in communication with each other.

@@

Match--girls were the only group in \citeyear{booth1903life} to have formed a trade union,
according to \citeauthor{booth1903life}'s account at the time~\cite{booth1903life}.
\citeauthor{10.2307/3827491} noted that match--girls
``\dots~pooled their resources to purchase their plumes and clothes~\dots~and expressed their solidarity through small [and major] strikes''~\cite{booth1903life}.
But they were also, as \citeauthor{10.2307/3827491} confesses, known for ``brashness, irregularity, low morality, and little education''~\cite{10.2307/3827491}.
These were workers who treasured their independence, but also fiercely protected one another, contributing to the common good.
Their ``brashness'' for instance may have detracted from their public image, but almost undoubtedly contributed to their sense of solidarity,
making their propensity to act against such unfair treatment and poor conditions understandable and maybe predictable.

\subsubsection{Industrial workers}
\topic{Piecework might be most familiar in the context of industrial and factory work, which largely defined manufacturing through the \nth{20} century.}
Before the factory assembly line arose, however, railway companies adopted piecework regimes in the early \nth{20} century.
What followed was a flourishing of piecework management practices,
as railway companies worked to find effective ways
to motivate and evaluate this skilled workforce of engineers.
\citeauthor{10.2307/23702539} takes up a case study of the Santa Fe Railway,
finding that they employed ``efficiency experts'' to develop a ``standard time''
to determine pay for each task at the company informed by
``thousands of individual operations''; % ~\cite{10.2307/23702539}.
\citeauthor{10.2307/23702539} goes on to list
some of the roles required to facilitate piecework
in the early \nth{20} century
--- among them, ``piecework clerks, inspectors, and `experts'~''~\cite{10.2307/23702539}.
This oversight, while controversial
(especially among workers~\cite{american1921problem}),
paved the way for piecework to grow substantially.

\topic{The 1930s represented a boom for piecework on an unprecedented scale,
especially among engineering and metalworking industries.}
\citeauthor{hart2013rise} characterize the 1930s
--- and more broadly the first half of the \nth{20} century ---
as the ``heyday'' of the use of piecework.
They attribute this to the shortage of male workers,
who would have gone through a conventional apprenticeship process
affording them more comprehensive knowledge of the total scope of work.

\topic{Piecework found its way into the war effort during World War~II.}
With the vast majority of men drafted into service,
factories found themselves turning to
a predominantly female workforce that had neither
the formal training nor
the years of apprenticeship experience
that conventional workers would have had.
Rather than attempting to train this new labor force in every aspect of industrial work,
these women were trained for individual tasks
and assigned to that task.
``Rosie the Riveter'',
an icon of \nth{20} century America who
represented empowerment and opportunity for women~\cite{honey1985creating},
was a pieceworker~\cite{davies2014origins}.

>>>>

\topic{The HCI community has used the term ``piecework'' to describe
myriad instantiations of on--demand labor, but
researchers have generally made this allusion in passing.}
Since we trace a much stronger parallel between
(historical) piecework and (contemporary) on--demand work,
a more comprehensive background on piecework will be useful.
Specifically,
first, we'll define ``piecework'' as researchers in its field understand it;
and second, we'll trace the rise and fall of piecework at a high level,
identifying key figures and ideas during this time.
This section is not intended to be comprehensive:
instead, it sets up the scaffolding necessary for
our later investigations of on--demand work's three questions:
complexity limits,
task decomposition, and
worker relationships.

@@

\topic{Aligning on--demand work with piecework requires an understanding of what piecework is.}
While it has had several definitions over the years,
we can trace a constellation of characteristics that recur throughout the literature.
We'll follow this research, collecting
descriptions,
examples, and
definitions,
to develop a sense of piecework.

\topic{Piecework's history traces back further perhaps than most would expect.}
\citeauthor{grier2013computers} describes the process astronomers adopted of hiring young boys
to calculate equations in order
to better--predict the trajectories of various celestial bodies in the night sky~\cite{grier2013computers}.
George Airy was perhaps the first to rigorously put piecework--style decomposition to work;
by breaking complex calculations into constituent parts, and
training young men to solve simple algebraic problems,
Airy could distribute work to many more people than could otherwise complete the full calculations.

\topic{Piecework may have started in the intellectual domain of astronomical calculations and projections,
but it found its foothold in manual labor.}
Piecework took hold in farm work~\cite{hughRaynbirdTaskWork}, % hold foothold
in textiles~\cite{restructuringPieceworkBaker,riisOtherSideLives},
on railroads~\cite{Brown01041990}, and 
elsewhere in manufacturing~\cite{10.2307/3827491}.
At the turn of the \nth{20} century, \citeauthor{taylor1914principles}
--- a mechanical engineer with a fascination in work efficiency ---
began studying and formalizing the decomposition, tracking, and management of tasks.
In \citeyear{taylor1914principles} he published \citetitle{taylor1914principles},
concretizing an idea that had nebulously been forming,
and which he had been working out himself, for years~\cite{taylor1914principles}.
Scientific Management (and Fordism) thrust piecework into higher gear, especially as
mass manufacturing and
a depleted wartime workforce forced industry to find new ways to eke out more production capacity.

@@

\topic{By \citeyear{hughRaynbirdTaskWork} we find
a concise definition of piecework
in \citeauthor{hughRaynbirdTaskWork}'s essay on piecework,
particularly driven toward encapsulating the manual labor of farmwork.}
He does this by contrasting two paradigms ---
``the chief difference lies between the day--labourer,
who receives a certain some of money~\dots~for his day's work,
and the task--labourer, whose earnings depend on the quantity of work done''~\cite{hughRaynbirdTaskWork}.
\citeauthor{10.2307/2338394} defines it through examples:
``payment is made for each hectare which is pronounced to be well ploughed~\dots~for each living foal got from a mare;~\dots~for each living calf got''~\cite{10.2307/2338394}.
This framing offers an intuitive sense of piecework;
``payment for results,'' as he calls it,
is not only common in practice, but
well--studied in labor economics as well~\cite{Figlio2007901,weitzman1976new,10.2307/3003414,BJIR:BJIR038}.

\topic{It's worth acknowledging that
``this distinction [between piece--rates and time--rates] was not completely clear--cut''~\cite{hart2013rise}.}
Employers implemented piece--rates in some aspects and
time--rates in others.
The Rowan premium system, for example,
essentially paid workers
a base rate for time plus
additional pay depending on output~\cite{rowan1901premium}.
As \citeauthor{rowan1901premium}'s premium system guaranteed an hourly rate
regardless of the worker's productive output
\textit{as well as} additional compensation tied to performance,
workers were
in some senses ``task--labourers'', but
in other senses ``day--labourers''.
This was just one of several alternatives to strict time-- and piece--rate renumeration paradigms.

@@

\topic{We see this sense of independence in myriad times, locales, and industries.}
\citeauthor{10.2307/3827491} offers a look into the lives and culture of ``match--girls''
--- teenage women who assembled matchsticks in the late \nth{19} century in London. % added that these women were in London since Chinmay critiqued that we're glossing over the labor advocacy movement that grew in Europe, which revealed that we completely left out the fact that this all happened in England.
\ali{Chinmay: but still not convinced; this is a post hoc ergo propter hoc argument}
Of interest was their reputation ``\dots~for generosity, independence, and protectiveness,
but also for brashness, irregularity, low morality, and little education''~\cite{10.2307/3827491}.
\citeauthor{10.2307/27508091} document piecework from 1850 through 1930 in Australia,
finding similar notions of independence and autonomy among piecework newspaper compositors:
``If a piece--work compositor~\dots~decided that he did not want to work on a particular day or night, the management recognised his right to put a `substitute' or `grass' compositor in his place''~\cite{10.2307/27508091}.
This sense of independence and autonomy appears to be a common thread of piecework.

\topic{Since workers could now choose their own schedule and style, there arose a discussion on how best to manage pieceworkers.}
This conversation came to regard workers antagonistically~\cite{roy1954efficiency}, a far cry from the earlier rhetoric on piecework, which promised that
piece workers would gladly work as diligently for as long as possible because
incentive--based pay rewards hard work~\cite{clark1908cotton}.

@@

\topic{Piecework's popularity in the United States and Europe fell almost as quickly as it had climbed.}
Between 1938 and 1942, the proportion of metal workers under piecework systems had climbed steeply from 11\% to 60\%~\cite{hart2005piecework}.
By 1961, that proportion dropped to 8\%~\cite{carlson1982time}.
\citeauthor{carlson1982time} details that, from 1973 to 1980, the holdouts of piecework
--- where more than 50\% worked under incentive wage plans ---
were principally in clothes--making (e.g. hosiery, footwear, and garments).
\citeauthor{hart2013rise} offer a number of explanations for the sudden demise of piecework.
% \ali{Should I add ``(especially in the United States but more broadly in the developed world)'' or something? Since a lot of developing nations still have sweatshops and piecework to some extent or another?}
The salient suggestions include:
\begin{inlinelist}
\item the emergence of more effective, more nuanced incentive models
--- rewarding teams for complex achievements, for instance;
\item the shifting of piecework industries such as manufacturing and textiles to other countries; and
\item the quality of ``multidimensional'' work, which was too difficult to evaluate~\cite{hart2013rise}.
\end{inlinelist}

@@

Viewing on--demand work as a modern instantiation of piecework is relatively straightforward by this definition.
First, platforms such as Amazon Mechanical Turk (AMT), Uber, Upwork, and TaskRabbit pay by the task, though some mix systems in similar ways to the Rowan system's combination of piece rate and time rate pay.
Second, workers are attracted to these platforms by the freedom they offer to pick the time and place of work~\cite{martin2014being,whyWouldAnyoneBrewer}.
Third, system developers as on Mechanical Turk typically assume no professional skills in transcription or other areas, and attempt to build that expertise into the work flow~\cite{noronha2011platemate,bernsteinSoylent}.
Given this alignment, many of the same properties of piecework historically will apply to on--demand work as well.

\subsection{Case studies in piecework}
\topic{Throughout the rest of the paper, we will return to four case studies to frame our analyses:
Airy's employment of human computers;
domestic and farm workers;
the ``match--girls'' strike;
and industrial and assembly-line workers.}
In introducing these cases at a high level,
we'll trace the history of piecework
while also framing the later analysis of the leading research threads we named earlier:
complexity, decomposition, and relationships.

@@

This practice ensured two outcomes that disfavored workers.
First, it eliminated any potential to advance professionally, as
workers' careers ended relatively quickly,
and without formal education in mathematics they struggled to find work for which their experience was meaningful.
Second, it limited workers' ability to organize
by ensuring that workers were in relatively little communication with each other.

@@

Match--girls were some of the earliest to have formed a trade union,
according to \citeauthor{booth1903life}'s account in \citeyear{booth1903life}. % ~\cite{booth1903life}.
\citeauthor{10.2307/3827491} noted that match--girls
``\dots~pooled their resources to purchase their plumes and clothes~\dots~and expressed their solidarity through small [and major] strikes''~\cite{booth1903life}.
But they were also, as \citeauthor{10.2307/3827491} confesses, known for ``brashness, irregularity, low morality, and little education''~\cite{10.2307/3827491}.
These were workers who treasured their independence, but also fiercely protected one another. % , contributing to the common good.
Their ``brashness'' for instance may have detracted from their public image, but almost undoubtedly contributed to their sense of solidarity,
making their propensity to act against such unfair treatment and poor conditions understandable and maybe predictable.

\subsubsection{Industrial workers}
\topic{Piecework might be most familiar in the context of industrial and factory work, which largely defined manufacturing through the \nth{20} century.}
Before the factory assembly line arose, however, railway companies adopted piecework regimes at the turn of the \nth{20} century.
What followed was a flourishing of management practices,
as railway companies worked to find effective ways
to motivate and evaluate this skilled workforce of engineers.
\citeauthor{10.2307/23702539} takes up a case study of the Santa Fe Railway,
finding that they employed ``efficiency experts'' to develop a ``standard time''
to determine pay for each task at the company informed by
``thousands of individual operations'';
\citeauthor{10.2307/23702539} goes on to list
some of the roles required to facilitate piecework
in the early \nth{20} century
--- among them, ``piecework clerks, inspectors, and `experts'~''~\cite{10.2307/23702539}.
This oversight, while controversial
(especially among workers~\cite{american1921problem}),
paved the way for piecework to grow substantially.

\topic{The 1930s represented a boom for piecework on an unprecedented scale,
especially among engineering and metalworking industries.}
\citeauthor{hart2013rise} characterize the 1930s
--- and more broadly the first half of the \nth{20} century ---
as the ``heyday'' of piecework.
They attribute this to the shortage of male workers,
who would have gone through a conventional apprenticeship process
affording them more comprehensive knowledge of the total scope of work.

\topic{Piecework found its way into the war effort during World War~II.}
With the vast majority of men drafted into service,
factories found themselves turning to
a mostly female workforce that had neither
the formal training nor
the years of apprenticeship experience
that conventional workers would have had.
Rather than attempting to train this new labor force in every aspect of industrial work,
these women were trained for individual tasks
and correspondingly assigned to that or a similar task.
``Rosie the Riveter'',
an icon of \nth{20} century America who
represented empowerment and opportunity for women~\cite{honey1985creating},
would have been a pieceworker~\cite{davies2014origins}. \ali{was? would have been?}
